---
title: "Adding data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  echo = FALSE
)
```

```{r setup}
library(appliedepidata)
```

# Guide to contributing new datasets

## PACKAGE FOLDER STRUCTURE

To below is a simplified explanation of the [R packages data
chapter](https://r-pkgs.org/data.html). For a fuller understanding, read
that chapter.

### Overview

``` bash
.
├── appliedepidata.Rproj
├── _pkgdown.yml
├── data
│   └── internaliseddata.rda
├── data-raw
│   └── newdata.R
├── inst
│   └── extdata
│       └── tableoftables.xlsx
│       └── newdata.xlsx
├── R
│   └── newdata_doc.R
└── man
    └── newdata.Rd
```

### Folders you will edit

-   **inst/extdata**: Newly added datasets go in here in their original
    filetype, e.g. .xlsx, .csv, but also .rds files. These files will be
    downloaded with the `save_data()` function.\
-   **data-raw**: Contains R scripts used for processing and
    internalizing datasets (making them more efficient for storage),
    i.e. for turning newly added datasets in `inst/extdata` to .rda
    files
-   **R**: Contains R scripts in Roxygen2 format which define the
    functions and datasets. When run this creates the R documentation
    files (.Rd files) which are saved in the **man** folder. Multiple
    datasets can be defined in one script, which is sensible as some are
    grouped together (e.g. part of the same outbreak)
-   **vignettes**: Contains detailed instructions on how to use
    particular functions. They complement the function-level
    documentation in the **man** folder by giving broader explanation.

### Folders you will not edit

These contain files outputted by running the code in the folders above.

-   **data**: Contains internalised and processed R datasets (.rda
    files).
-   **man**: Contains the R documentation files for package functions
    and datasets (created when running the code in the **R** folder).
    Each .Rd file has information about a specific function or dataset.
-   **tests**: Contains test files for developer use only.
-   **internal data**: When you build a package, the `Rda` datasets
    (from `data` folder) can become "internal" (more efficient for file
    storage). These are accessed by calling `package::dataset` (e.g.
    `appliedepidata::AJS_AmTiman`. They can also be imported directly
    from github using link to the file in data folder e.g.
    'rio(<github-raw url>)', or the `appliedepidata::get_data` or
    `appliedepidata::save_data` functions.
-   **sysdata**: Not relevant for current package setup. In some setups
    you are supposed to put *tableoftables* in sysdata (i.e. just for
    package usage). However for our current setup leave in *extdata*.

## ADDING DATA

This describes the process for adding a file to the repo. Note that the
processes for adding a non-R file (any file that is not `.rda`) and an R
file (any file already in`.rda` format) are slightly different.

If you are adding a dataset from an existing R package, you can skip to
step 3 below.

1.  **Name your file appropriately**

    A. You can name it whatever you want, but stick to basic naming
    conventions.

    B. Ensure that there is not already file in *tableoftables.xlsx*
    named the same.

    C. Avoid generic names like: `linelist_cleaned.xlsx` or
    `survey_data.xlsx`.

    D. Use consistent and descriptive names without spaces (e.g.,
    `AJS_AmTiman`, `sitrep_mortality_survey`).

    E. Name files from the same group (e.g. from the same outbreak or
    for the same case study) with the same prefix. E.g:
    `examplename_data` and `examplename_population` for a case linelist
    and corresponding denominator table respectively.

2.  **Place your file in the correct folder**

    A. A *non-rda* file (e.g. `xlsx`, `shp`, `zip`, `rds`) goes in
    `inst/extdata` folder. This is what will be downloaded when running
    the save_data() function. i. If adding a shapefile then zip it

    B. An *rda* file goes in `data` folder

3.  **Build the package with the added data**

    A. Press Ctrl + Shift + B to build the package with the newly added
    data. This will mean the data is recognized for the next step.

<!-- -->

4.  **Reproducibly edit dataset and internalize** (see
    `data-raw/AJS_AmTiman.R` for example)

    A. In your console run
    `usethis::use_data_raw("<name of your file without extension>")`.

    B. This creates an R script in the `data-raw` folder. It will
    already contain a comment at the top saying "code to prepare
    <dataset name> data goes here" and a `usethis::use_data()` function
    to internalise the dataset (i.e. to produce the rda file)

    C. Edit the R script to correctly read in the file using
    `system.file()`, as below. Necessary edits to the dataset should
    also go here. Then run this script to internalise the data, and
    close.

```{r, eval=FALSE}
  
## code to prepare `examplename_linelist` dataset goes here

# Define the path to the Excel file in inst/extdata
file_path <- system.file("extdata", "<examplename_linelist.xlsx>", package = "appliedepidata")

# Read in the Excel file using rio
examplename_linelist <- rio::import(file_path)

# Other code for editing the file can go here




# Save the data as an internal .rda file in the data/ directory
usethis::use_data(examplename_linelist, overwrite = TRUE)


```

5.  **Add the dataset to the `tablesoftables.xlsx` as described below.**

6.  **Add documentation for each dataset added**

    A. Create a new R script in the `R` folder, with one file for all
    datasets in a group. The file name should be the group name
    generated from tableoftables (column P), with suffix '\_doc'. Note
    the easiest is to copy and edit an existing R script in the folder.

    B. Ensure to clearly document the source and license for the
    dataset.

    C. Ensure to put the correct name of the dataset at the bottom of
    each dataset description (under @docType). E.g.
    `examplename_linelist`.

    D. Run devtools::document() to create the actual R documentation.
    This will be an .Rd file in the `man` folder, with file name
    corresponding with the dataset name (e.g. `examplename_linelist.Rd`)

<!-- -->

6.  **Add the datasets to `_pkgdown.yml`**

    A. Subtitle: Describe the group of linelists. State the year in
    brackets and language at the end

    B. Contents: List the datasets. Again, the names here correspond to
    the name of the dataset without extension, e.g
    "examplename_linelist".

    C. Make sure to have correct indentation and use of dashes. See
    prior examples in the file.

## ADDING DATASET METADATA (adding to `tablesoftables.xlsx`)

Below is a table explaining how to fill in each variable in the dataset
metadata Excel sheet (`tablesoftables.xlsx`). This guide helps ensure
consistency and completeness when adding new datasets to your
collection.

-   **name**: The filename of the dataset as it appears in the
    `inst/extdata` directory, **without** the file extension. This
    should be unique within the dataset group, and ideally also within
    the *tableoftables* (i.e. avoid generic names like:
    `linelist_cleaned.xlsx` or `survey_data.xlsx`). Use consistent and
    descriptive names without spaces (e.g., `AJS_AmTiman`,
    `mortality_survey`).

-   **type**: The category or type of the dataset (e.g., `linelist`,
    `population`, `shape`, `survey`, `dictionary`).

-   **extension**: The file extension (e.g., `xlsx`, `zip`).

-   **type_version**: Used to identify the *original* dataset and its
    associated child data. Increment when format or variables change. If
    there are multiple linelists in one group, this would increment with
    the type.

-   **data_version**: Used to identify the *original* dataset and its
    associated child data. Increment when format or variables change.
    Ensure you document changes in the appropriate 'data-raw' file.

-   **language**: Language code using [ISO 639-1
    codes](https://en.wikipedia.org/wiki/List_of_ISO_639-1_codes) (e.g.,
    `en`, `fr`).

-   **country**: Country code using [ISO 3166-1 alpha-3
    codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) (e.g.,
    `tcd`).

-   **scale**: Geographic scale (e.g., `subnational`, `national`, or
    `international`).

-   **subject**: Main subject of the dataset (e.g.,
    `acute jaundice syndrome`).

-   **context**: Context of the data (e.g., `outbreak`, `survey`).

-   **fictional**: Is the dataset fictional (`yes`) or real (`no`)?

-   **year**: Year the data was collected (e.g., `2016`). This is the
    *earliest* year in the dataset (even if fictional).

-   **description**: Brief description of the dataset. Ideally, copy
    from roxygen documentation.

-   **usage**: Intended usage (e.g., `{sitrep} walkthroughs`,
    `training`).

-   **license**: License for dataset (e.g., `gpl3`, `mit`).

-   **group_identifier**: *DO NOT EDIT* - Created by concatinating
    function in excel. High-level identifier combining `subject`,
    `context`, `country`, and `year`
    (e.g.,`acute_jaundice_syndrome_outbreak_tcd_2016`).

-   **unique_identifier**: *DO NOT EDIT* - Combines `group_identifier`,
    `type`, `type_version`, `data_version`, `context`, and `year` to
    create a unique identifier (e.g.
    `acute_jaundice_syndrome_outbreak_tcd_2016_linelist_1`).

For example, when adding an Ebola dataset, you would enter the
information as shown below. The original dataset (whether it's from
{outbreaks} or another source) would be considered `type_version` 1. If
it's the only linelist in its group, it remains `type_version` 1. If a
completely different linelist is added (not just an edited version),
increment the `type_version` accordingly.

For any changes to the data (such as cleaning or changing nums of rows
or columns), increment the `data_version` (e.g., `data_version` 2), but
the `type_version` remains the same to indicate that it's a derivative
(or "child") of the original. Each child dataset gets its own entry.

If a dataset is translated into a different language, create a new entry
for the translated version while keeping the `data_version` and
`type_version` the same, but editing the `language` column accordingly.
This ensures you can trace back the parent-child relationship between
datasets.

| **Variable**          | **Example Entry**                                    |
|----------------------------|--------------------------------------------|
| **name**              | `ebola_linelist_cleaned`                             |
| **type**              | `linelist`                                           |
| **extension**         | `xlsx`                                               |
| **type_version**      | `1`                                                  |
| **data_version**      | `1`                                                  |
| **language**          | `en`                                                 |
| **country**           | `lbr`                                                |
| **scale**             | `national`                                           |
| **subject**           | `ebola`                                              |
| **context**           | `outbreak`                                           |
| **fictional**         | `yes`                                                |
| **year**              | `2014`                                               |
| **description**       | Linelist data from the Ebola virus                   |
|                       | disease outbreak in Liberia in                       |
|                       | 2014\.                                               |
| **usage**             | `introexercises`, etc.                               |
| **license**           | `gpl3`                                               |
| **group_identifier**  | `ebola_outbreak_lbr_2014`                            |
| **unique_identifier** | `ebola_outbreak_lbr_2014_linelist_1_1_outbreak_2014` |
