---
title: "Adding data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{adding-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  echo = FALSE
)
```

```{r setup}
library(appliedepidata)
```



# Adding a new dataset to the package  

- guideline for contributing datasets similar to [tidytuesday](https://github.com/rfordatascience/tidytuesday/blob/master/.github/pr_instructions.md)


# repo folder structure 

Read this [r packages chapter](https://r-pkgs.org/data.html) for details. 


```bash
.
├── appliedepidata.Rproj
├── data
│   └── newdata.rda
├── data-raw
│   └── newdata.R
├── inst
│   └── extdata
│       └── tableoftables.xlsx
│       └── newdata.xlsx
├── R
│   └── newdata_doc.R
└── man
    └── newdata.Rd
```



- folder structure 
  - **data** R datasets go in data folder
  - **inst/extdata**Non-R datasets go in inst>extdata
    - option for adding messy datasets. However this would then not have 
    an internalised .rda version as described below.  
  - **internal data** additional complication is that when you build a package, 
  you can make the Rda datasets (from data folder) "internal" (more efficient 
  for file storage, as then become part of binary), and these are then access 
  by doing package::dataset. They can also be imported directly from github 
  using link to the file in data folder e.g. 'rio(<github-raw url>)'. 
  - **data-raw** usually contains R scripts which are used for creating the 
  exported or internal data (e.g. if have edited dataset, or where {usethis}
  internalises dataset)
  - **sysdata** probably the *tableoftables* should just stay in *extdata* but
  alternatively could go in sysdata which is not exported (i.e. just for package
  usage) 


## if a non-R (not .rda) file

- name file what want but stick to guidelines described in metadata below
- versioning of datasets (and edits thereof, e.g. same data for diff 
bits of course)
- define overarching source file and resulting child files (as per metadata below)
- place file in 'inst/extdata' folder
 - if adding a shapefile then zip it 
- in console run 'usethis::use_data_raw(<name of your file w/out extension>)
- read in the file with 'rio' and 'system.file'
- make edits as needed
- save with usethis::usedata()
- add documentation for each data set added (ideally group these within the 
same file if part of the same group of datasets)
 - ideally: data dictionary used to fill in the *man* files for each (link to 
 function)
 - probably do the same way [alison horst](https://github.com/lter/lterdatasampler/) does
- licensing: while the overall package repo will be GPL3, it is possible that individual datasets will come under a different license (so there needs to be a license section in documentation for each dataset)

    
## if adding an .rda file: 

- link to the appropriate part from non-R onwards where relevant



  

# Data Entry Guide for Dataset Metadata

Below is a table explaining how to fill in each variable in the dataset 
metadata Excel sheet (1tablesoftables.xlsx). This guide helps ensure 
consistency and completeness when adding new datasets to your collection.

- **name**:  The filename of the dataset as it appears in the `inst/extdata`
  directory,  **without** the file extension. This should be unique within 
  the dataset group, and ideally also within the *tableoftables* 
  (i.e. avoid generic names like: `linelist_cleaned.xlsx` or `survey_data.xlsx`). 
  Use consistent and descriptive names without spaces (e.g., `AJS_AmTiman`,
  `mortality_survey`).

- **type**: The category or type of the dataset (e.g., `linelist`, 
  `population`, `shape`, `survey`, `dictionary`).

- **extension**: The file extension (e.g., `xlsx`, `zip`).

- **type_version**: Used to identify the *original* dataset and its associated
  child data. Increment when format or variables change. If there are multiple
  linelists in one group, this would increment with the type.

- **data_version**: Used to identify the *original* dataset and its associated 
  child data. Increment when format or variables change. Ensure you document 
  changes in the appropriate 'data-raw' file.

- **language**: Language code using [ISO 639-1 codes](https://en.wikipedia.org/wiki/List_of_ISO_639-1_codes) (e.g., `en`, `fr`).

- **country**:  Country code using [ISO 3166-1 alpha-3 codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) (e.g., `tcd`).

- **scale**: Geographic scale (e.g., `subnational`, `national`, or `international`).

- **subject**: Main subject of the dataset (e.g., `acute jaundice syndrome`).

- **context**: Context of the data (e.g., `outbreak`, `survey`).

- **fictional**:  Is the dataset fictional (`yes`) or real (`no`)?

- **year**: Year the data was collected (e.g., `2016`). This is the *earliest* 
  year in the dataset.

- **description**: Brief description of the dataset. Ideally, copy from 
  roxygen documentation.

- **usage**: Intended usage (e.g., `{sitrep} walkthroughs`, `training`).

- **license**: License for dataset (e.g., `gpl3`, `mit`).

- **group_identifier**: *DO NOT EDIT* - Created by concatinating function in excel.
  High-level identifier combining `subject`, `context`, `country`, and `year`
  (e.g.,`acute_jaundice_syndrome_outbreak_tcd_2016`).

- **unique_identifier**: *DO NOT EDIT* - Combines `group_identifier`, `type`,
  `type_version`, `data_version`, `context`, and `year` to create a unique 
  identifier (e.g. `acute_jaundice_syndrome_outbreak_tcd_2016_linelist_1`).


For example, when adding an Ebola dataset, you would enter the information as 
shown below. The original dataset (whether it's from {outbreaks} or another source)
would be considered `type_version` 1. If it’s the only linelist in its group, it 
remains `type_version` 1. If a completely different linelist is added (not just an
edited version), increment the `type_version` accordingly.

For any changes to the data (such as cleaning or changing nums of rows or columns),
increment the `data_version` (e.g., `data_version` 2), but the `type_version` 
remains the same to indicate that it’s a derivative (or "child") of the original.
Each child dataset gets its own entry.

If a dataset is translated into a different language, create a new entry for
the translated version while keeping the `data_version` and `type_version` the same, 
but editing the `language` column accordingly. 
This ensures you can trace back the parent-child relationship between datasets.


| **Variable**          | **Example Entry**                  |
|-----------------------|------------------------------------|
| **name**              | `ebola_linelist_cleaned`           |
| **type**              | `linelist`                         |
| **extension**         | `xlsx`                             |
| **type_version**      | `1`                                |
| **data_version**      | `1`                                |
| **language**          | `en`                               |
| **country**           | `lbr`                              |
| **scale**             | `national`                         |
| **subject**           | `ebola`                            |
| **context**           | `outbreak`                         |
| **fictional**         | `yes`                              |
| **year**              | `2014`                             |
| **description**       | Linelist data from the Ebola virus |
|                       | disease outbreak in Liberia in     |
|                       | 2014.                              |
| **usage**             | `introexercises`, etc.             |
| **license**           | `gpl3`                             |
| **group_identifier**  | `ebola_outbreak_lbr_2014`          |
| **unique_identifier** | `ebola_outbreak_lbr_2014_linelist_1_1_outbreak_2014`|

